---
title: "EDPS 845 Assignment 1"
author: "Jayden Nord"
---

## Overview

Demonstrate your mastery of concepts and procedures covered in *An Introduction to R*, chapters 1 through 9. See Appendix A for examples.

Complete this assignment by responding to each question below using your own original code, in addition to answering any questions. Follow Google's [R coding conventions](https://google.github.io/styleguide/Rguide.xml), and use [R markdown](http://rmarkdown.rstudio.com/index.html) formatting for your responses, with example R code written as

```{r}
# Comment within some example code
```

Your file should have an Rmd extension, and you should check that it compiles to HTML using the *Knit* command in RStudio (this runs `knitr::knit()` on the file). Note that you can already *Knit* this file to see how it works.

## Exercises

### Chapter 1

1. Show that you can access the home page for the HTML R help menu, as well as help files for individual functions.

```{r, eval=FALSE}
# launches home page for HTML R help menu
help.start()

# get individual help files
help("summary")
?summary # equivalent to help("summary")
```

2. Demonstrate correct and incorrect naming of R objects.

```{r, eval = FALSE}
# correct naming conventions
somename    
some_name   
some_name_02

# incorrect naming conventions
02_some_name
some name   

# ill advised naming conventions
some.name
```

3. How do you list all available objects in the R workspace? And how do you remove individual objects?

```{r}
# make some objects
x <- 10; y <- 5; z <- 2

# ls() lists all available objects in workspace
ls() 

# remove objects with rm()
rm(x, y)    # removes object x and y
ls()        # confirm removal
```

### Chapter 2

1. Create and index a vector using a logical, integer, negative integer, and character vector.

```{r}
# some vectors
pets <- c("dogs", "cats", "birds", "hamsters", "mice", "snakes")
pets2 <- c(20, 30, 15, 50, 20, 19) # just some random numbers
names(pets2) <- pets

# logical index
pets[c(TRUE, TRUE, TRUE, FALSE, FALSE, TRUE)]

# integer index
pets[c(1:3, 6)]

# negative integer index
pets[-c(4, 5)]

# character index
pets2[c("dogs", "cats", "birds", "snakes")]
```

2. Create a simple data frame containing a numeric variable, character variable, and factor with at least 3 levels. Use the data frame to demonstrate how R codes and handles missing data. 

```{r}
# make data frame
d <- data.frame(
  x = 1:10,
  y = letters[1:10],
  z = factor(rep(1:3, length.out = 10), 
             labels = c("Dog", "Cat", "Bird")),
  stringsAsFactors = FALSE)
print(d)

# demonstrate how R handles missing data
# make some data missing
d[2, "x"] <- NA
d[5, "y"] <- NA
d[3, "z"] <- NA
print(d)
```


### Chapter 3

1. Demonstrate how to check and change the modes and attributes of R objects.

```{r}
x <- 1:10
y <- letters[1:10]

# check mode
mode(x); mode(y)

# change mode
z <- as.character(x)
mode(z)
```


### Chapter 4

1. Complete the exercise on page 17, where you calculate 95% confidence intervals for state mean incomes.

```{r}
state <- c("tas", "sa", "qld", "nsw", "nsw", "nt", "wa", "wa",
  "qld", "vic", "nsw", "vic", "qld", "qld", "sa", "tas",
  "sa", "nt", "wa", "vic", "qld", "nsw", "nsw", "wa",
  "sa", "act", "nsw", "vic", "vic", "act")

statef <- factor(state)

incomes <- c(60, 49, 40, 61, 64, 60, 59, 54, 62, 69, 70, 42, 56,
  61, 61, 61, 58, 51, 48, 65, 49, 49, 41, 48, 52, 46,
  59, 46, 58, 43)

# I go a little overboard here...
funs <- c("mean", "length", "var")
desc <- data.frame(sapply(funs, function(f) tapply(incomes, statef, f)))
desc <- within(desc, {
  error <- qt(.975, df = length - 1) * sqrt(var / length)
  lower <- mean - error
  upper <- mean + error
})
```

```{r}
desc[, c("lower", "mean", "upper")]
```

2. In what order are the levels of ordered factors stored in?

Levels of ordered factors are stored in alphabetical order.


### Chapter 5

1. Create an array with dimensions `c(3, 6, 2)` and use it to demonstrate array indexing.

```{r}
arr <- array(1:(3 * 6 * 2), dim = c(3, 6, 2))
arr
arr[1, ,]
arr[1, 3, ]
arr[1, 3, 2]
arr[ , 3, ]
```

2. Create a two dimensional array of 100 integers sampled with replacement from the vector 1:20, and use an index array to set the diagonal values to 0.
```{r}
elements <- sample(1:20, size = 100, replace = TRUE)
mat1 <- elements
dim(mat1) <- c(10, 10)
i1 <-c(1:10, 1:10)
dim(i1) <- c(10, 2)
mat1[i1] <- 0
mat1
```

3. Construct the same array using the `array()` function and using a vector with the `dim()` set. Prove they are the same.

```{r}
mat2 <- array(elements, dim = c(10, 10))
i2 <- array(c(1:10, 1:10), dim = c(10, 2))
mat2[i2] <- 0
mat2

# test whether mat1 and mat2 are the same
identical(mat1, mat2)
```

4. Use your array from 5.1, and another with the same dimensions, to demonstrate how arithmetic works with arrays.

```{r}
arr2_elements <- sample(1:50, size = 3 * 6 * 2, replace = TRUE)
arr2 <- array(arr2_elements, dim = c(3, 6, 2))

arr + arr2
arr * arr2
```


5. Show how to clear the dimensions of an array.

```{r}
vec <- c(mat1) # clears dimensions of mat1
vec
```


### Chapter 6

1. Demonstrate the different ways of subsetting a list.

```{r, eval = FALSE}
d <- data.frame(matrix(1:9, ncol = 3))
colnames(d) <- c("dog", "cat", "bird")

# the following are equivalent
d$dog
d[, "dog"]

# more complex subsetting
d[1:2, c("cat", "bird")]
d[d$dog > 1, "dog"]
```


2. What is the difference between giving multiple lists to `list()` versus `c()`?

Lists via `list()` are comprised of "components" which can be either individual values or entire objects of varying dimensions. There may be a mixture of different types of components of different classes. List components can also have names.

Concatenations via `c()` are combinations of values of vectors into larger vectors. The resulting vector inherits the class of the concatenated values or coerces the elements into a single class. Concatenations cannot be a mixture of different data types (i.e. numeric and character).

```{r, error=TRUE}
# example list
example_list <- list(animal = "dog", feeding_times = c(3, 7, 9))
str(example_list)
example_list

# example concatenation
example_c <- c(5, 1:10, "dog")
str(example_c)
example_c
```

3. How do data frames in R relate to lists?

Data frames are a special case of lists. Data frames are lists of vectors, factors, numeric matricies, lists, and/or other data frames. The vectors in a data frame must have the same length. Matrices in a data frame must have the same row length.

4. What is the purpose of `attach()`?

Consider some list `example_list` with components `animal` and `feeding_times`. If one wanted to access just `animal`, one would use the code `example_list$animal`. It may be tedious to type the list name if repeatedly accessing components. 

`attach()`, with the list name as an argument, bypasses the need to type `$` to access components. For example, with...
```{r, eval = FALSE}
example_list <- list(animal = "dog", feeding_times = c(3, 7, 9))
```

This block...
```{r, eval = FALSE}
example_list$animal
example_list$feeding_times
```

...is equivalent to this block.
```{r, eval = FALSE}
attach(example_list)
animal
feeding_times
detach(example_list)
```


### Chapter 7

1. Demonstrate the different approaches to reading in data from an external file.

```{r}
# make a data file
cat("animal number fed", "dog 3 2", "cat 12 4",
    sep = "\n",
    file = "some_data.dat")

# read in white-space delimited file "some_data.dat" which has a header
d <- read.table("some_data.dat", header = TRUE)
d

# read data into a vector or list of vectors
# scan() will read in elements separated by spaces or carriage returns
#  into a character vector
d <- scan("some_data.dat", what = character())
d

# assuming "some_data.dat" has a character vector and 2 numeric vectors
# the line below will function similarly to d <- read.table but skips 
#  the in-file line with the header
d <- scan("some_data.dat", what = list("", 0, 0), skip = 1)
d

# remove the data file
unlink("some_data.dat")
```


### Chapter 8

1. Demonstrate the use of distribution functions to simulate data and calculate probabilities and quantiles.

```{r}
# simulate 10 cases from normal distribution
#   with pop. mean of 1 and pop. sd of 1.33
rnorm(n = 10, mean = 1, sd = 1.33)

# calculate quantile from normal distribution given 
#  mean = 1
#  sd = 1.33
#  probability = .05
qnorm(p = .05, mean = 1, sd = 1.33)

# calculate probability from normal distribution given
#  mean = 1
#  sd = 1.33
#  quantile = 1.187655
pnorm(q = -1.187655, mean = 1, sd = 1.33)

# more useful example:
# get p-value from a two-tailed t-statistic of 1.45 with df = 87
pt(q = 1.45, df = 87)

# get absolute value of t-critical given alpha = .05 and df = 87
qt(p = 1 - .05/2, df = 87)
```
  
2. Simulate data from a t distribution and check their normality.

```{r}
# simulated data
d <- rt(n = 20, df = 87)
d

# check normality
qqnorm(d)
```

### Chapter 9

1. What is the difference between `if() else` and `ifelse()`?

The `ifelse()` function is a vectorized version of `if() else`. The function `ifelse(condition, a, b)` returns some vector with elements:
* `a[i]` where `condition[i] == TRUE` and 
* `b[i]` where `condition[i] == FALSE`

For example:
```{r}
x <- sample(40:60)
y <- ifelse(x > 50, 10, 5)
print(data.frame(x = x, condition = x > 50, y = y))
```

2. Demonstrate the use of a `for()` loop.

```{r}
for(i in 1:8){
  cat("Iteration", i, "\n")
}
```

