---
title: "EDPS 845 Assignment 2"
author: "Jayden Nord"
---

## Exercises

### Data structures
1. What are the six types of atomic vector? How does a list differ from an atomic vector?

Logical, integer, double, character, complex, and raw.

2. What makes `is.vector()` and `is.numeric()` fundamentally different to `is.list()` and `is.character()`?

The functions `is.vector()` and `is.numeric()` are general mode tests. That is, `is.vector` returns `TRUE` for any mode that only has name attributes while `is.numeric()` will return `TRUE` for vectors of mode integer or double. 

Meanwhile, the functions `is.list()` and `is.character()` are specific mode tests. They will return true if the object is specifically a list and specifically a character, respecitvely.

3. Test your knowledge of vector coercion rules by predicting the output of the following uses of `c()`:

    ```{r, eval=FALSE}
    c(1, FALSE)   
    c("a", 1)     
    c(list(1), "a")
    c(TRUE, 1L)   
    ```
    
```{r, collapse = TRUE}
c(1, FALSE)       # numeric vector of 1 0
c("a", 1)         # character vector of "a" "1"
c(list(1), "a")   # list of length 2 with a numeric value and character value
c(TRUE, 1L)       # numeric vector of 1 1
```

4. Why do you need to use `unlist()` to convert a list to an atomic vector? Why doesn't `as.vector()` work? 

According to the help file for `as.vector()`, the function `as.vector()` does not coerce to atomic vectors but coerces to vectors of some given mode. List is a valid mode, so `as.vector()` coerces a list to a list. The function `unlist()` specifically flattens a list into an atomic vector.

5. Why is `1 == "1"` true? Why is `-1 < FALSE` true? Why is `"one" < 2` false?

* `1 == "1"` is true because the numeric 1 is coerced to a character.
* `-1 < FALSE` is true because the logical FALSE is coerced to a numeric which turns out to be 0. 
* `"one" < 2` is false because first, the numeric 2 is coerced to a character. Certain characters are "greater" than others. `"one"` begins with the letter "o" which is greater than the character 2.

6. Why is the default missing value, `NA`, a logical vector? What's special
   about logical vectors? (Hint: think about `c(FALSE, NA_character_)`.)

There is no clear indication in the text. However, logical tends to be the bottom of the coercion hierarchy. `NA` will be coerced up to the appropriate mode depending on other content in the vector. If `NA` was of some higher mode, any intended vector elements will be coerced up to match `NA` when it should be the other way around (as illustrated in the hint).

7. An early draft used this code to illustrate `structure()`:

    ```{r, collapse = TRUE}
    structure(1:5, comment = "my attribute")
    ```

    But when you print that object you don't see the comment attribute. Why? Is the attribute missing, or is there something else special about it? (Hint: try using help.)

By default, only object values are printed. The associated attributes are not printed but are still a part of the object. This fact can be confirmed with the `str()` or `attributes`  functions. For example:

```{r, collapse = TRUE}
x <- structure(1:5, comment = "my attribute")

str(x)

attributes(x)
```

8. What happens to a factor when you modify its levels? 
    
    ```{r, results = "none"}
    f1 <- factor(letters)
    levels(f1) <- rev(levels(f1))
    ```
    
The labels for the factors are substituted with the new set of labels. The underlying data is unchanged. To illustrate:

```{r, collapse = TRUE}
f1 <- factor(letters)
str(f1)

levels(f1) <- rev(levels(f1))
str(f1)
```

9. What does this code do? How do `f2` and `f3` differ from `f1`?

    ```{r, eval = FALSE}
    f2 <- rev(factor(letters))
    
    f3 <- factor(letters, levels = rev(letters))
    ```

`f2` is the reverse vector of `f1` but has the same levels as `f1`. `f2` prints the alphabet backwards while `f1` prints the alphabet forwards.

`f3` is the reverse vector of `f1` and has the reverse levels of `f1`. `f1` and `f3` print the alphabet forwards.

10. What does `dim()` return when applied to a vector?

`NULL` is returned.

11.  If `is.matrix(x)` is `TRUE`, what will `is.array(x)` return?

`is.array(x)` will also return `TRUE` because matrices are merely special cases of arrays. If something is a matrix, then by definition that something is an array.

12. How would you describe the following three objects? What makes them different to `1:5`?

    ```{r}
    x1 <- array(1:5, c(1, 1, 5))
    x2 <- array(1:5, c(1, 5, 1))
    x3 <- array(1:5, c(5, 1, 1))
    ```

The set of `x`s are arrays of mode numeric with three dimensions, two of which have length 1 and the other of length 5. The vector `1:5` has only one dimension of length 5.

13. What attributes does a data frame possess?

Data frames possess names, row names, and class attributes. For example.

```{r, collapse=TRUE}
attributes(mtcars)
```

14.  What does `as.matrix()` do when applied to a data frame with 
    columns of different types?

The function will attempt to coerce the different column types to the same mode as whatever column has the mode highest on the coercion hierarchy. 

15.  Can you have a data frame with 0 rows? What about 0 columns?

Yes and yes!

```{r, collapse = TRUE}
d <- data.frame()
str(d)
```

### Subsetting

1.  Fix each of the following common data frame subsetting errors:

    ```{r, eval = FALSE}
    mtcars[mtcars$cyl = 4, ]
    mtcars[-1:4, ]
    mtcars[mtcars$cyl <= 5]
    mtcars[mtcars$cyl == 4 | 6, ]
    ```
    
```{r, eval = FALSE}
# corrected
mtcars[mtcars$cyl == 4, ]
mtcars[-c(1:4), ]
mtcars[mtcars$cyl <= 5, ]
mtcars[mtcars$cyl == 4 | mtcars$cyl == 6, ]
```

2.  Why does `x <- 1:5; x[NA]` yield five missing values? (Hint: why is 
    it different from `x[NA_real_]`?)

The logical vector of `NA` is recycled to the same length as `x`. A missing values in the resulting logicl index will always return missing, hence the five missing values. 

3.  What does `upper.tri()` return? How does subsetting a matrix with it 
    work? Do we need any additional subsetting rules to describe its behaviour?

    ```{r, eval = FALSE}
    x <- outer(1:5, 1:5, FUN = "*")
    x[upper.tri(x)]
    ```

The function `upper.tri()` returns a numeric vector of the upper triangle of the matrix `x`, read column-by-column. Subsetting with `upper.tri()` is subsetting with an integer index. The matrix `x` is first laid out as an atomic vector column-by-column. Then the integer index is used to return values which, here, happen to equal the index values.

4.  Why does `mtcars[1:20]` return an error? How does it differ from the 
    similar `mtcars[1:20, ]`?
    
An error is returned because mtcars has more than one dimension. An attempt is being made to index on only one dimension. `mtcars[1:20, ]` correctly recognizes the 2 dimensions and will resolve.

5.  Implement your own function that extracts the diagonal entries from a
    matrix (it should behave like `diag(x)` where `x` is a matrix).
    
```{r, collapse = TRUE}
x <- outer(1:5, 1:5, FUN = "*")

own_diag <- function(x){
  length <- nrow(x)
  if(length != ncol(x)) stop("Not a square matrix!")
  i <- cbind(1:length, 1:length)
  return(x[i])
}

own_diag(x)

identical(diag(x), own_diag(x))
```

6.  What does `df[is.na(df)] <- 0` do? How does it work?

Wherever there are `NA` values in `df`, those values are made to equal 0. The `is.na(df)` function builds a logical array of same dimensions as `df`. Values of `TRUE` are wherever there are `NA` values. 0's are assigned to the dimensional indices on `df` corresponding to wherever the logical array is `TRUE`.

7.  Given a linear model, e.g., `mod <- lm(mpg ~ wt, data = mtcars)`, extract the residual degrees of freedom. Extract the R squared from the model summary (`summary(mod)`)
    
```{r, collapse=TRUE}
mod <- lm(mpg ~ wt, data = mtcars)

mod$residuals

summary(mod)$r.squared
```

8.  How would you randomly permute the columns of a data frame? (This is an important technique in random forests.) Can you simultaneously permute the rows and columns in one step?

Columns (and rows, simultaneously) can be randomly permuted by randomly sampling the dimension names into a new order and using that random sample as an index for the dimension. For example:

```{r, collapse = TRUE}
# original mtcars
head(mtcars)

# random permuted columns and rows
head(mtcars[sample(rownames(mtcars)), sample(colnames(mtcars))])
```

9. How would you select a random sample of `m` rows from a data frame? What if the sample had to be contiguous (i.e., with an initial row, a final row, and every row in between)?

```{r, collapse = TRUE}
# example data frame
d <- data.frame(x = 1:10, y = 11:20)

# random sample of rows
m <- sample(rownames(d))
m

# contiguous random sample of rows
rows <- rownames(d)
last <- length(rows)
between <- rows[(1 + 1):(last - 1)]
m <- c(rows[1], sample(between), rows[last])
m
```

10.  How could you put the columns in a data frame in alphabetical order?

By using order on the column names of the data frame in order to get the index of the column names in ascending order.

```{r, collapse = TRUE}
ordered <- order(colnames(mtcars))
head(mtcars[, ordered])
```

### Vocabulary

There are no exercises for this chapter.

### Style

There are no exercises for this chapter.

### Functions

1.  What function allows you to tell if an object is a function? What function allows you to tell if a function is a primitive function?
    
```{r, collapse = TRUE}
# is.function() tells that an object is a function

x <- 5

is.function(mean)

is.function(x)

# is.primitive() tells that a function is a primitive function

is.primitive(sum)
```

2.  This code makes a list of all functions in the base package. 
    
    ```{r}
    objs <- mget(ls("package:base"), inherits = TRUE)
    funs <- Filter(is.function, objs)
    ```

    Use it to answer the following questions:

    a. Which base function has the most arguments?
    
```{r, collapse = TRUE}
num_formals <- sapply(funs, function(i) length(formals(i)))
most_args <- which.max(num_formals)
names(most_args)
```
    
    b. How many base functions have no arguments? What's special about those
       functions?
       
```{r, collapse = TRUE}
no_formals <- which(num_formals == 0)

length(no_formals)

# Most are primitive functions while others make calls to internal functions

# A primitive function...
names(no_formals[45])
as.double

# A function that calls an internal...
names(no_formals[100])
geterrmessage
```
       
    c. How could you adapt the code to find all primitive functions?
    
```{r, collapse = TRUE}
# Before, had funs <- Filter(is.function, objs)
prims <- Filter(is.primitive, objs)
length(prims)
```

3. What are the three important components of a function?

    1. body, the code that makes up the function
    2. formals, also known as the arguments
    3. environment, location of the function's variables

```{r, collapse = TRUE}
some_function <- function(x) {
  deviates <- (x - mean(x))
  return(deviates^2)
}

body(some_function)

formals(some_function)

environment(some_function)
```

4. When does printing a function not show what environment it was created in?

Whenever a function is a primitive, it will not show what environment it was created in. For example: 

```{r, collapse = TRUE}
print(sum)
```

5. What does the following code return? Why? What does each of the three `c`'s mean?

    ```{r, eval = FALSE}
    c <- 10
    c(c = c)
    ```

The above code returns the value 10 with name "c". In `c(c=c)`, the first `c` is the concatenate function with argument `c = c`. The second `c` or the `c` immediately left of the equal sign is a name to be associated with whatever is to the right of the equal sign. The last `c` is the object `c` which holds the numeric value 10.

6. What are the four principles that govern how R looks for values?

The four principles are:

Principle | Description 
-------------------| -------------------------------------------
name masking             | If the function cannot find the name of an object defined within the function, it will look for it one level above.
functions vs. variables  | Like name masking; involves searching for function names than other object names.
a fresh start            | Each time a function is run, an environment is created for that specfic invocation of that function. Value definitions in a function occur within that freshly created environment. 
dynamic lookup           | Values are searched only at each invocation of the function.

7. What does the following function return? Make a prediction before running the code yourself.

    ```{r, eval = FALSE}
    f <- function(x) {
      f <- function(x) {
        f <- function(x) {
          x ^ 2
        }
        f(x) + 1
      }
      f(x) * 2
    }
    f(10)
    ```

The above will return 202. Essentially, the function does `((10 ^ 2) + 1) * 2` which equals 202.

8.  Clarify the following list of odd function calls:

    ```{r, eval = FALSE}
    x <- sample(replace = TRUE, 20, x = c(1:10, NA))
    y <- runif(min = 0, max = 1, 20)
    cor(m = "k", y = y, u = "p", x = x)
    ```

```{r, eval = FALSE}
# clarified function calls
x <- sample(c(1:10, NA), size = 20, replace = TRUE)
y <- runif(n = 20, min = 0, max = 1)
cor(x, y, method = "kendall")
```

9.  What does this function return? Why? Which principle does it illustrate?
  
    ```{r, eval = FALSE}    
    f1 <- function(x = {y <- 1; 2}, y = 0) {
      x + y
    }
    f1()
    ```

The above function returns the value 3. This demonstrates the "a fresh start" principle. The default for `x` assigns a value to `y`, overwriting the default option of 0. 

10.  What does this function return? Why? Which principle does it illustrate?

    ```{r, eval = FALSE}
    f2 <- function(x = z) {
      z <- 100
      x
    }
    f2()
    ```

The above function returns 100. This demonstrates the "dynamic lookup" principle. The value of `x` is not searched until it's called. Within the function, `z` is assigned the value of 100. Now that `z` exists within the function, calling `x` will call an existing `z` and thus the value of `z`.

11. Create a list of all the replacement functions found in the base package. 
   Which ones are primitive functions?

```{r, collapse = TRUE}
objs <- mget(ls("package:base"), inherits = TRUE)
objs_names <- names(objs)

# replacement functions
grep(pattern = "<-", objs_names, value = TRUE)

prims <- Filter(is.primitive, objs)
prims_names <- names(prims)

# primitive replacement functions
grep(pattern = "<-", prims_names, value = TRUE)
```

12. What are valid names for user-created infix functions?

User-created infix functions must start and end with `%`. Any character or string can be between the `%`s. 

13. Create an infix `xor()` operator.

```{r, collapse = TRUE}
x <- c(TRUE, TRUE, FALSE)
y <- c(TRUE, FALSE, FALSE)
'%xor%' <- function(x, y) xor(x, y)

xor(x, y)

x %xor% y
```

14. Create infix versions of the set functions `intersect()`, `union()`, and `setdiff()`.

```{r, collapse = TRUE}
'%i%' <- function(x, y) intersect(x, y)
'%u%' <- function(x, y) union(x, y)
'%d%' <- function(x, y) setdiff(x, y)

# make data based on help file from help(union)
x <- sample(1:20, 9)
y <- sample(3:23, 7)

x
y

# demonstrate infix functions
x %i% y
x %u% y
x %d% y
```

15. Create a replacement function that modifies a random location in a vector.

```{r, collapse = TRUE}
x <- letters
x

'rando<-' <- function(x, value){
  i <- sample(1:length(x), size = 1)
  x[i] <- value
  x
}

rando(x) <- "modified"
x
```

16. How does the `chdir` parameter of `source()` compare to `in_dir()`? Why might you prefer one approach to the other?

Both `source()` and `in_dir()` go to some directory to run some code. The default parameter for `chdir` is `FALSE`, meaning R's working directory will not be changed to the directory wherever the code is run. The function `source()` is meant to run code contained in text files. The function `in_dir()` does not access text files to run code.

17. What function undoes the action of `library()`? How do you save and restore the values of `options()` and `par()`?

Packages are unloaded by using `detach("package:package_name")` where `package_name` is the name of the package one wants to unload. For example:

```{r, collapse = TRUE}
# load the epmr package 
library(epmr)

# unload the epmr package
detach(package:epmr)
```

Values of of `options()` and `par()` can be saved by running the function and assigning the result to an object. The values can be restored by using the new object as an argument in the original function. For example:

```{r, eval = FALSE}
# save par values
saved_par <- par()

# restore par values
par(saved_par)

# save options values
saved_ops <- options()

# restore options values
options(saved_ops)
```

18. Write a function that opens a graphics device, runs the supplied code, and closes the graphics device (always, regardless of whether or not the plotting code worked).

```{r, eval = FALSE}
x11()              # opens graphics device
plot(1:10, 1:10)   # plots some supplied code
dev.off()          # closes graphics device
```

19.  We can use `on.exit()` to implement a simple version of `capture.output()`.

    ```{r, eval = FALSE}
    capture.output2 <- function(code) {
      temp <- tempfile()
      on.exit(file.remove(temp), add = TRUE)

      sink(temp)
      on.exit(sink(), add = TRUE)

      force(code)
      readLines(temp)
    }
    capture.output2(cat("a", "b", "c", sep = "\n"))
    ```

    Compare `capture.output()` to `capture.output2()`. How do the functions 
    differ? What features have I removed to make the key ideas easier to see? 
    How have I rewritten the key ideas to be easier to understand?

The original `capture.output()` has additional arguments and its body has more code with various `if` statements. The function `capture.output2` makes clear the functionality of the `on.exit` command which evaluates its given arguments after the function has completed. In the function, a temp file is created to which results from `code` are sinked to and also read to the console. At the end of the function the temporary file is deleted and sinking is returned to the console.